---
title: "COVID-19 Data Visualizations"
output: html_document
---

## Setup

```{r warning=FALSE, message=FALSE}
library(devtools)
library(ggplot2)
library(tidyr)
library(readr)
library(purrr)
library(dplyr)
library(stringr)
library(forcats)
library(rclipboard)
library(shiny)
```

## Data

The dataset is created by an Economist Joachim Gassen as a Github Package "tidycovid19" to facilitate the direct download of various Covid-19 related data from authoritative sources. The main source is Johns Hopkins University CSSE Github Repo. It also provides a flexible function and accompanying shiny app to visualize the spreading of the virus. 

Blog Link: https://www.r-bloggers.com/meet-tidycovid19-yet-another-covid-19-related-r-package/ 

Github Link: https://github.com/joachim-gassen/tidycovid19

```{r message=FALSE}
install_github("joachim-gassen/tidycovid19")

library(tidycovid19)
covid19_dta <- download_merged_data(silent = TRUE, cached = TRUE)
```

```{r}
df <- covid19_dta %>%
  select(country, iso3c, date, confirmed, deaths, recovered, region, income, population, land_area_skm, pop_density, pop_largest_city, life_expectancy, gdp_capita, lockdown, timestamp)
```


## Visualizations

We can use the  function plot_covid19_spread to visualize the four values of interest; confirmed cases, deaths, recoveries, and active cases (confirmed - recovered). Particularly, I set the y-axis to be the number of cases per 100,000 people in these countries to compare the numbers across the countries.

Documentation: https://rdrr.io/github/joachim-gassen/tidycovid19/man/plot_covid19_spread.html

```{r}
plot_covid19_spread(
  type = "confirmed",
  df, 
  intervention = "lockdown",
  highlight = c("ITA", "ESP", "FRA", "DEU", "USA","CHN","BRA","GBR", "BEL"),
  edate_cutoff = 50,
  per_capita = TRUE, 
  per_capita_x_axis = TRUE,
  population_cutoff = 10,
  min_cases = 0.1
)

plot_covid19_spread(
  type = "deaths",
  df, 
  intervention = "lockdown",
  highlight = c("ITA", "ESP", "FRA", "DEU", "USA","CHN","BRA","GBR", "BEL"),
  edate_cutoff = 50,
  per_capita = TRUE, 
  per_capita_x_axis = TRUE,
  population_cutoff = 10,
  min_cases = 0.1
)

plot_covid19_spread(
  type = "recovered",
  df, 
  intervention = "lockdown",
  highlight = c("ITA", "ESP", "FRA", "DEU", "USA","CHN","BRA","GBR", "BEL"),
  edate_cutoff = 50,
  per_capita = TRUE, 
  per_capita_x_axis = TRUE,
  population_cutoff = 10,
  min_cases = 0.1
)

plot_covid19_spread(
  type = "active",
  df, 
  intervention = "lockdown",
  highlight = c("ITA", "ESP", "FRA", "DEU", "USA","CHN","BRA","GBR", "BEL"),
  edate_cutoff = 50,
  per_capita = TRUE, 
  per_capita_x_axis = TRUE,
  population_cutoff = 10,
  min_cases = 0.1
)
```


## Interactive Shiny App

If you find these graphs helpful, Mr. Gassen actually created a shiny app so you can chage the setting and make your own plots. 

Link: https://jgassen.shinyapps.io/tidycovid19/


